---
  output: html_document
editor_options: 
  chunk_output_type: console
---
  
  # elementary entities
  
  # dplyr pipeline 
  
  ```{r library}
library(dplyr) # grammar data manipulation 
```

```{r data importation}
carabids <- read.delim ('https://raw.githubusercontent.com/zdealveindy/anadat-r/master/data/carabid-beetles-boreal-forest.txt', row.names = 1)
data("iris")
```

*Use of %>% pipe (magrittr, implemented in dplyr)*
  ```{r %>% pipe}
iris$Species %>% summary()
summary(iris$Species)
# 1) susbet species
iris_setosa1 <- filter(
  .data = iris, 
  Species=="setosa") 
# 2) ration petal length/ petal width 
iris_setosa1 <- mutate(.data = iris_setosa1, ratio_pet_lengthwidth = Petal.Length/Petal.Width)
# 1) and 2) in a single step
iris_setosa2 <- filter(
  .data = iris, 
  Species=="setosa") %>% 
  mutate(ratio_pet_lengthwidth = Petal.Length/Petal.Width) %>%
  arrange(desc(by = ratio_pet_lengthwidth)) 
# subseting setosa species, creating a ratio column and arrange it by decreasing values 
```
When using %>%, do not need to notify the 1st argument which correspond to the object we are creating  
--
  (f-g)(x) = f(g(x))
x -> g(x) -> f(g(x))
--
  Here we have: 
  x = iris
y = setosa 
g() = filter()
f() = mutate()
h() = arrange()
i() = desc()
as:
  y <- g(x) -> f(g(x)) -> h(f(g((i(x))))) 
-- 
  we are taking x, going through different functions (one by one) to get to y object

*Use of %in%*
  1) to compare sequences of numbers:
  a %in% b
2) to test if value is in column 
1.3 %in% Petal.Length
3) iris %>% filter(species %in% species2)
```{r %in%}
#1 use
## to check if vectors match
a <- seq(-2,10, by = 1.5) %>% print()
b <- seq(-2,5) %>% print()
a %in% b
which(a %in% b, arr.ind = T) # a in b -> will give us values based on a order 
which(b %in% a, arr.ind = T)
## or if a value is inside a vector
0.5 %in% iris$Sepal.Length
# 2 use
# Creating a dataframe:
B <- subset(iris, Species == "versicolor" | Species == "virginica")
B <- subset(iris, Species != c("setosa"))
B <- subset(iris, Species %in% c("setosa","virginica"))  
dataf1 <- data.frame(Type = c("Fruit","Fruit","Fruit","Fruit","Fruit",
                              "Vegetable", "Vegetable", "Vegetable", "Vegetable", "Fruit"),
                     Name = c("Red Apple","Strawberries","Orange","Watermelon","Papaya",
                              "Carrot","Tomato","Chili","Cucumber", "Green Apple"),
                     Color = c(NA, "Red", "Orange", "Red", "Green",
                               "Orange", "Red", "Red", "Green", "Green"))
# Adding a New Column:
dataf2 <- within(dataf1, {
  Red_Fruit = "No"
  Red_Fruit[Type %in% c("Fruit")] = "No"
  Red_Fruit[Type %in% "Vegetable"] = "No"
  Red_Fruit[Name %in% c("Red Apple", "Strawberries", "Watermelon", "Chili", "Tomato")] = "Yes"
})
```

*{ } Curly braces* 
  Curly braces are used to evaluate a series of expressions (separated by new lines or semicolons) and return only the last expression:
  ```{r}
f <- function() {x <- 1; y <- 2; x + y}
f()
ff <- function() {
  xx <- 2
  yy <- 4
  xx + yy
}
ff()
```

*"&"; "&&"; "|"; "!"; *
  "&" (ie and): vectorized, they can return vector
"&&" (ie and): not vectorized (can not return values)
"|" (ie or)
```{r}
m <- -2:2 %>% print()
and_v <- (m >= 0) & (m <= 0)
or_v  <- (m >= 0) | (m <= 0) 
and_nv <- (m >= 0) && (m <= 0)
or_nv  <- (m >= 0) || (m <= 0) 
```


*Use with if()*
  ```{r}
if (test_expression) {
  statement
}
# want to create a new column "anthropic affinity based on the number of individuals counted in each habitat
## we consider "resistant" species the one with > 30 occurences
### with IF 
if (carabids$managed.forests > 30) {
  carabids1 <- dplyr::mutate(.data = carabids, 
                             anthropic_spp = carabids$managed.forests > 30)
} 
carabids1$anthropic_spp <- as.factor(carabids1$anthropic_spp) %>% 
  plyr::revalue(c("TRUE" = "resistant",
                  "FALSE" = "sensitive"))
class(carabids1$anthropic_spp)
### with ifelse
carabids2 <- ifelse(test = carabids$managed.forests > 30,
                    yes = "resistant",
                    no = "sensitive") %>% 
  cbind(carabids)
colnames(carabids2)[which(colnames(carabids2) == ".")] <- "anthropic_spp"
class(carabids2$anthropic_spp)
### multiple conditions IF
#### simple
if (carabids$moist.primeval.forests > 30 & carabids$dry.primeval.forests > 25) 
{ carabids3 <- dplyr::mutate(.data = carabids, 
                             primeval_spp = c(carabids$moist.primeval.forests > 30 & carabids$dry.primeval.forests > 25))
}  # we got what we wanted even if warning message
```

*IF & ELSE*
  A, B, C are dataframes.

B exists and A doesn't exist      -> C = B
A exists and B does not exist     -> C = A
A & B exist                       -> C = A + B 
A & B do not exist                -> C = NULL

```{r if and else}
A <- subset(iris, Species == "setosa")
B <- subset(iris, Species == "versicolor" | Species == "virginica")
B <- subset(iris, Species != c("setosa"))
B <- subset(iris, Species %in% c("setosa","virginica"))          
            
ABC_fct <- function() {
  if (!exists("A") & exists("B")) {
    C= B
} else if (exists("A") & !exists("B")) {
    C= A
} else if (exists("A") & exists("B")) {
    C= rbind(B,A)
} else {C <- NULL}
}
ABC_fct() %>% summary()
rm(A)
ABC_fct() %>% summary()
rm(B)
ABC_fct() %>% summary()
```

*square brackets []* 
brackets allows to get inside an object
if 2 dimensions: [rows, columns]
```{r}
numberlist <- c(3, 5, 7, 5, 3, 2, 6, 8, 5, 6, 9)
numberlist[c(2)] # second element
numberlist[-1] # everything except 1st 
numberlist[-c(1:5)] # everything except 5 1sts
numberlist[numberlist > 8 | numberlist <= 3] # multiple condition selection
iris[,which(colnames(iris)=="Species")] %>% summary() # complicated version; good if you do not know the column number and if you want to avoid $
iris$Species %>% summary()
summary(iris$Species)
# [[]] to suppress the element name
carabids3[6]
carabids3[[6]]
carabids3[,6] # same as [[]]
carabids3[c(1:5),6]
```