---
title: "CLEANING_TRANSFO_HALASEN"
output: html_document
date: "2023-04-12"
---

*NOTES* TOASK : ctrl + F this and search - new variable : upper - lower - resolve otu pb - *LIBRARY*

```{r packages}
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(dplyr) # A Grammar of Data Manipulation
library(tidyverse) # A Grammar of Data Manipulation
library(vegan) # ecological statistical analysis
library(phyloseq) # Handling and analysis of high-throughput microbiome census data 
library(forcats) # Tools for Working with Categorical Variables
library(MiscMetabar) # Analyze and visualize metabarcoding data
library(circlize) # circular visualization
library(ggVennDiagram) # venn diagrams 
library(phyloseq.extended) # phyloseq packages with more options
library (microbiomeutilities) # for rarefaction curves
library(multcompView) # Visualizations of Paired Comparisons
library(plyr)
library(pals)
library(ade4)
library(lme4)
library(lmtest)
library(reshape2)
library(data.table)
library(viridis) # palettes and themes
library(hrbrthemes) # palettes and themes
```
```{r color palette}
pals::pal.bands(alphabet, alphabet2, cols25, glasbey, kelly, polychrome, 
  stepped, tol, watlington,
  show.names=FALSE)

funky <- grDevices::colorRampPalette(c("#A6CEE3","#1F78B4","#B2DF8A",
"#33A02C","#FB9A99","#E31A1C",
"#FDBF6F","#FF7F00","#CAB2D6",
"#6A3D9A","#FFFF99","#B15928"))
```
```{r functions}
outersect <- function(x, y) {
  sort(c(setdiff(x, y),
         setdiff(y, x)))
} # extract uncommon values between x - y 

stde <- function (x) {sqrt(var(x))/length(x)}
```
```{r dummy calc}
# hellinger : sum of reads per sample = 1 ? (as same as relative abundance ?)

#abund1 <- c(sample(0:250, n = 10))
```

*DATA IMPORTATION AND MODIF*

```{r raw_data}
setwd("H:/REDACTION/RESULTS/cutting_ecm/TRANSF_DATA2") 
#data run
run_data <- read.table("run_data.txt", header=T, sep="\t",na.strings=T,dec=".")
colnames(run_data)

# all clusters
raw_clust <- read.table("raw_clusters.txt",header=T,sep="\t",na.strings=T,dec=".") %>%
  arrange(desc(Cluster_Size)) # importing and arrange by decreasing cluster size
raw_clust[raw_clust==""] <- NA # putting NA when empty
raw_clust[,4] <- as.character(raw_clust[,4]) # reference as.character
raw_clust$Cluster_ID <- gsub('scata5670_','c',raw_clust$Cluster_ID) # replace character occurence in all cells
rownames(raw_clust) <- raw_clust$Cluster_ID
order <- raw_clust$Cluster_ID 
order <- gsub('c','',raw_clust$Cluster_ID) # creating a vector order so we can arrange later
raw_clust <- data.frame(cbind(order, raw_clust))
raw_clust[,"order"] <- as.numeric(raw_clust[,"order"])

# subsetting fungi X non-fungi
raw_f_clust <- subset(raw_clust,!grepl("PLANT|NF",raw_clust[,5]))
raw_nf_clust <- subset(raw_clust,grepl("PLANT|NF",raw_clust[,5]))
```
```{r tax-eco table }
# only the 1st 1000 clusters 
taxeco_id <- read.table("tax_table.txt",header=T,sep="\t") # change #N/A in excel in NA
length(taxeco_id) # number of columns 
length(rownames(taxeco_id)) # number of rows (ie clusters, 1st 1000 fungi)
taxeco_id[,c(3:length(taxeco_id))] <- lapply(taxeco_id[,c(3:length(taxeco_id))], as.factor) 
rownames(taxeco_id) <- taxeco_id$Cluster_ID # rownames of corresponding clusters
taxeco_0 <- subset(taxeco_id[,c(3:length(taxeco_id)),]) 

taxeco2 <- subset(taxeco_id[,c(3:length(taxeco_id)),]) 
taxeco2 <- matrix(as.character(unlist(taxeco2)),
              ncol = ncol(taxeco2),
              nrow = nrow(taxeco2))
```

```{r subsetting taxtable}
id_taxeco <- raw_f_clust %>%
  filter(Cluster_ID %in% taxeco_id$Cluster_ID) %>%
  arrange(order) # filtering taxtable identified manually
id_taxeco <- cbind("MANUAL_ID", id_taxeco)

unid_taxeco <- raw_f_clust %>%
  filter(!Cluster_ID %in% taxeco_id$Cluster_ID) %>%
  arrange(order) # filtering taxtable not identified 

taxeco_f <- dplyr::bind_rows(id_taxeco, unid_taxeco)
colnames(taxeco_f)[1] <- "ID_TYPE"
taxeco_f$ID_TYPE[is.na(taxeco_f$ID_TYPE)] <- "AUTO_ID" 

taxeco_id <- cbind("MANUAL_ID", taxeco_id)
colnames(taxeco_id)[1] <- "ID_TYPE"

# not same length row between id_taxeco (998) and taxeco (999): not normal, can cause offset
match_id <- taxeco_id$Cluster_ID %in% id_taxeco$Cluster_ID
as.factor(match_id) %>% 
  levels()
which(match_id == "FALSE", arr.ind = T) # mismatch at row 240

taxeco_id <- taxeco_id[-240,]

match_id <- taxeco_id$Cluster_ID %in% id_taxeco$Cluster_ID # checking if resolved
as.factor(match_id) %>% 
  levels()
which(match_id == "FALSE", arr.ind = T) # good

###
taxeco_id$Cluster_Size <- id_taxeco$Cluster_Size # reattributing the proper cluster_size values

taxeco1 <- merge(taxeco_f, taxeco_id, all = T) %>%
  arrange(order) # merging taxeco and idtaxeco 

rownames(taxeco1) <- taxeco1$Cluster_ID

# putting reference of AUTO_ID in OTU_ID
which(taxeco1[,"ID_TYPE"] == "MANUAL_ID") %>% length() # finding the rows NOT concerned
taxeco1$OTU_ID <- as.character(taxeco1$OTU_ID) # putting as character for next line
taxeco1$OTU_ID <- c(taxeco1$OTU_ID[1:998],  taxeco1$Reference[999:length(taxeco1$Reference)]) # merging reference and otu_id 
  
# subsetting interesting variables
colnames(taxeco1)
taxeco <- taxeco1[, c(17:24,1)]
taxeco <- sapply(taxeco, as.character) 
rownames(taxeco) <- rownames(taxeco1) # cluster name in row
```

```{r sample data}
# sample data
sam <- read.table("sample_table.txt",header=T,sep="\t",na.strings = T) 
sam_no_match <- read.table("samples_no_match.txt",header=T,sep="\t") 

sam[sam==""] <- NA # filling the blanck by NAs
rownames(sam) <- sam$sample_id # after transposing, resolving pb with headers

# subsetting design samples per site 
sam_hala <- subset(sam, sam$site == "Halasen") # 61 samples
sam_controls_na <- filter(sam, !site %in% c("Halasen", "Fangamon", "Halasen_p_5")) # 19 samples (or controls)
```

```{r otu table}
otu_all <- read.table("otu_table.txt",header=F,sep="\t",na.strings = T) 
otu_all = setNames(data.frame(t(otu_all[,-1])), otu_all[,1]) # transpose + 1st column (ie samples set as header
## samples = columns, otu abundance = rows
colnames(otu_all)[1] <- "Cluster_ID" # after transposing, resolving pb with headers
nrow(otu_all) # total number of all clusters (ie fungi ecm or no-ecm, plants)
rownames(otu_all) <- otu_all$Cluster_ID
otu_all <- otu_all[-1] # deleting 1st column (cluster_id is now in rownames)
colnames(otu_all)
cluster_names_otu_all <- rownames(otu_all)

# removing clusters with 0 rowsums 
## relics of PCR mismatch tag (still amplified)
## but clusters still there after the exclusion of mismatch tags (eg ITS7-49 x ITS4-48)
otu_all <- sapply(otu_all, as.numeric) %>% as.data.frame()
rownames(otu_all) <- cluster_names_otu_all
which(rowSums(otu_all) == 0) %>% length() # clust with 0 reads = 557 (4050 - 557 = 3493)
otu <- otu_all %>%
  filter(!rowSums(otu_all) ==0) # suppression of these 557 clusters from otu matrix
nrow(otu)
ncol(otu)

# controls and samples with 0 exclusion 
not_samples_kept <- c("sITS-standard2", "sITS-standard1", "sPCR_neg_1", "sPCR_neg_2", "sQ2", "s49", "s97") # cf MM_RESULTS.Rmd ```{r DESIGN SAMPLES}  to understand how they are found 

samples_kept_pq <- otu[, colnames(otu) %in% not_samples_kept] 
clusters_in_controls <- which(samples_kept_pq >0, arr.ind = T) %>% 
  rownames() %>% as.factor() %>% levels() %>% print() # 13 clusters found in controls samples
otu_samples_out <- otu[, !(colnames(otu) %in% not_samples_kept)] # suppressing these samples from otu matrix
ncol(otu_samples_out)

# excluding NF clusters to rarefy only with fungi
otu_fungi <- otu_samples_out %>%
  filter(rownames(otu_samples_out) %in% rownames(taxeco)) # taxeco has only fungi clusters

# creating otu_fungi_hala only for hala to rarefy on the unique same site
otu_fungi_t <- t(otu_fungi) %>% as.data.frame()
otu_fungi_t <- otu_fungi_t %>%
  filter(rownames(otu_fungi_t) %in% rownames(sam_hala)) 
otu_fungi <- t(otu_fungi_t) # transposing back
nrow(otu_fungi) # 3493 - 3411 = 82 clusters suppressed

# transfo into matrix
otu1 <- matrix(as.numeric(unlist(otu_fungi)),
              ncol = ncol(otu_fungi),
              nrow = nrow(otu_fungi)) # puting array numeric and transforming in matrix for further functions 

# re-attribution (after being deleted from matrix transfo)
rownames(otu1) <- rownames(otu_fungi)
colnames(otu1) <- colnames(otu_fungi)
```

```{r otu <1% suppresion}
# suppression of <1 %
otu_relative <- prop.table(otu1, margin = 2) # margin = 2 -> column sum and relative abundances
verif_sum <- colSums(otu_relative) %>% summary() %>% print() # # verification: ok if 1 and some NAs
otu_001_transf <- ifelse(otu_relative>0.01,otu1,0) # creating a new matrix where <1% reads are transformed into 0
otu_001 <- subset(otu_001_transf, !rowSums(otu_001_transf)==0)

nrow(otu_001_transf)
nrow(otu_001)
nrow(otu_001_transf) - nrow(otu_001) # 2822 clusters with relative abundance < 1% = suppressed -> 589 conserved 

otu_notrarefied <- otu_001

# data properties BEFORE - AFTER suppresion of <1%
rd_sps_inf1 <- data.frame(colnames(otu1), colSums(otu1), colSums(otu_001)) # read sum per sample
colnames(rd_sps_inf1) <- c("sample_id", "sumreads_otu1", "sumreads_otu_001")
summary(colSums(otu1))
summary(colSums(otu_001))
str(rd_sps_inf1)

rd_sps_inf1_gg <- reshape2::melt(rd_sps_inf1[,2:3])
ggplot(rd_sps_inf1_gg, aes(x = value, color = variable, fill = variable))+
  geom_histogram(aes(y = ..count..), alpha = 0.4, position = "identity")+
  geom_density(alpha = 0.3)+
  scale_y_continuous(name = "nb of samples")+
  scale_x_continuous(name = "read sum per sample", breaks = seq(0,3000 ,by = 200))+
  ggtitle("Number of samples per read sum")+
  geom_vline(xintercept = c(mean(rd_sps_inf1$sumreads_otu1), mean(rd_sps_inf1$sumreads_otu_001)),lwd = 0.6,colour = c("red","blue"))
```

```{r hellinger transformation}
otu_hellinger <- sapply(otu_relative, sqrt)
otu_hellinger <- matrix(as.numeric(unlist(otu_hellinger)),
              ncol = ncol(otu_001_transf),
              nrow = nrow(otu_001_transf))
rownames(otu_hellinger) <- rownames(otu_001_transf)
colnames(otu_hellinger) <- colnames(otu_001_transf)
verif_sum_hell <- colSums(otu_hellinger) %>% summary() %>% print() # # verification: ok if 1 and some NAs
```

```{r otu rarefaction}
# data properties BEFORE rarefaction
rd_sps <- data.frame(rownames(otu_t), rowSums(otu_t)) # read sum per sample
hist(rd_sps$rowSums.otu_t.)

# rarefaction based on all 
otu_t0 <- t(otu_001)
otu_t <- matrix(as.numeric(unlist(otu_t0)),
              ncol = ncol(otu_t0),
              nrow = nrow(otu_t0))
rownames(otu_t) <- rownames(otu_t0)
colnames(otu_t) <- colnames(otu_t0)

otu_rarefied <- as.data.frame(rrarefy(otu_t, min(rowSums(otu_t)))) # apply rarefaction
otu_rarefied <- t(otu_rarefied)
otu_rarefied <- matrix(as.numeric(unlist(otu_rarefied)),
              ncol = ncol(otu_rarefied),
              nrow = nrow(otu_rarefied))

rownames(otu_rarefied) <- rownames(otu_001)
colnames(otu_rarefied) <- colnames(otu_001)

# data properties BEFORE - AFTER rarefaction
summary(colSums(otu_001))
rd_sps_raref <- data.frame(colnames(otu_001), colSums(otu_001), colSums(otu_rarefied)) # read sum per sample
colnames(rd_sps_raref) <- c("sample_id", "sumreads_notrarefied", "sumreads_otu_rarefied")
summary(colSums(otu_001))
summary(colSums(otu_rarefied))
str(rd_sps_raref)

rd_sps_raref_gg <- reshape2::melt(rd_sps_raref[,2:3])
ggplot(rd_sps_raref_gg, aes(x = value, color = variable, fill = variable))+
  geom_histogram(aes(y = ..count..), alpha = 0.4, position = "identity")+
  geom_density(alpha = 0.3)+
  scale_y_continuous(name = "nb of samples")+
  scale_x_continuous(name = "read sum per sample", breaks = seq(0,3000 ,by = 200))+
  ggtitle("Number of samples per read sum") # raref worked good
```

*CREATION OF PHYSEQ OBJECTS*
```{r physeq objects}
# creating the physeq objects 
otu_phy_rarefied <- otu_table(otu_rarefied, taxa_are_rows = TRUE)
otu_phy_notrarefied <- otu_table(otu_notrarefied, taxa_are_rows = TRUE)
otu_phy_hellinger <- otu_table(otu_hellinger, taxa_are_rows = TRUE)
otu_phy_relative <- otu_table(otu_relative, taxa_are_rows = TRUE)
otu_phy_001kept <- otu_table(otu1, taxa_are_rows = TRUE)
taxeco_phy <- phyloseq::tax_table(taxeco)
sam_phy <- sample_data(sam)

# general physeq object
physeq <- phyloseq(otu_phy_rarefied,taxeco_phy,sam_phy) %>% 
clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
physeq_rarefied <- physeq
physeq_notrarefied <- phyloseq(otu_phy_notrarefied,taxeco_phy,sam_phy) %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
# 2822 taxa removed from the <1% relative abundance suppresion. 

physeq_hellinger <- phyloseq(otu_phy_hellinger,taxeco_phy,sam_phy) %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
physeq_relative <- phyloseq(otu_phy_relative,taxeco_phy,sam_phy) %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
physeq_001kept <- phyloseq(otu_phy_001kept,taxeco_phy,sam_phy) 
# 10 taxa = erasing taxa from the control samples (among 13, 3 of them are probably also in others informative samples)
clusters_in_controls # controls to understand

# verif sample N
sample_names(physeq)
length(sample_names(physeq)) # 154 samples, as expected 
```

```{r taxa attribution mistakes}
# reattribution of correct names
physeq@tax_table[physeq@tax_table == "Gautieria_"] <- "Gautieria"
physeq@tax_table[physeq@tax_table == "Piloderma__"] <- "Piloderma"
physeq@tax_table[physeq@tax_table == "Sebacina__"] <- "Sebacina"
physeq@tax_table[physeq@tax_table == "Inocybe_geophylla"] <- "Inocybe_geophylla_(coll.)"

# changing variable levels
levels(as.factor(physeq@tax_table[,7])) # primary lifestyle levels

physeq@tax_table[,7] <- plyr::revalue(physeq@tax_table[,7],
                                      c("algal_parasite" = "other_lifestyle",
                                        "animal_parasite" = "other_lifestyle",
                                        "dung_saprotroph" = "saprotroph",
                                        "ectomycorrhizal" = "ectomycorrhizal",
                                        "lichen_parasite" = "other_lifestyle",
                                        "lichenized" = "other_lifestyle",
                                        "litter_saprotroph" = "saprotroph",
                                        "mycoparasite" = "other_lifestyle",
                                        "plant_pathogen" = "other_lifestyle",
                                        "root_endophyte" = "other_lifestyle",
                                        "soil_saprotroph" = "saprotroph",
                                        "sooty_mold" = "other_lifestyle",
                                        "unspecified_saprotroph" = "saprotroph",
                                        "wood_saprotroph" = "saprotroph"))
```
```{r transforming factor labels}
# hala
physeq@sam_data
## transforming specification into treatment factor
print(physeq@sam_data)
factor(physeq@sam_data$specification) %>% levels()
physeq@sam_data$treatment_edge <- plyr::revalue(physeq@sam_data$specification,
                           c("edge_north"="edge", 
                             "edge_south"="edge",
                             "kant_norra"="edge", 
                             "middle_lower"="middle", 
                             "middle_upper"="middle",
                             "midle"="middle"))
as.factor(physeq@sam_data$treatment_edge) %>% droplevels() %>% levels()

physeq@sam_data$specification_value <- plyr::revalue(physeq@sam_data$specification,
                           c("edge_north"="N", 
                             "edge_south"="S",
                             "kant_norra"="N", 
                             "middle_lower"="LO", 
                             "middle_upper"="UP",
                             "midle"= NA,
                             "middle"= NA))

physeq@sam_data <- within(physeq@sam_data, treatment_edge[treatment_edge=="middle" & treatment=="forest"] <- "FOR") 

physeq@sam_data$treatment_edge <- plyr::revalue(physeq@sam_data$treatment_edge,
                           c("middle"="CC_MID", 
                             "edge"="CC_EDG"))
as.factor(physeq@sam_data$treatment_edge) %>% droplevels() %>% levels()

# variables stream (upper or lower) and orientation (north or south)

physeq@sam_data$stream <- NA
physeq@sam_data$stream[which(physeq@sam_data$specification_value == "UP")] <- "UP"
physeq@sam_data$stream[which(physeq@sam_data$specification_value == "LO")] <- "LO"

physeq@sam_data$orient <- NA
physeq@sam_data$orient[which(physeq@sam_data$specification_value == "N")] <- "N"
physeq@sam_data$orient[which(physeq@sam_data$specification_value == "S")] <- "S"

phy <- physeq
```

```{r}
ecm_phy <- subset_taxa(physeq, primary_lifestyle == "ectomycorrhizal") %>% 
clean_pq(remove_empty_samples = T, remove_empty_taxa = T, clean_samples_names = F)
sapr_phy <- subset_taxa(physeq, primary_lifestyle == "saprotroph") %>% 
clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = F)
other_phy <- subset_taxa(physeq, (primary_lifestyle == "other_lifestyle" | is.na(primary_lifestyle)))  %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = F)
```

```{r subsets}
# subsetting based on sites and guilds 
# CTRL F + remove_empty_samples = T <-> remove_empty_samples = F

ecm_12 <- subset_samples(ecm_phy, ecm_phy@sam_data$year=="2012") %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
ecm_18 <- subset_samples(ecm_phy, ecm_phy@sam_data$year=="2018") %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)

orient <- subset_samples(ecm_phy, ecm_phy@sam_data$specification_value %in% c("N", "S")) %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)
stream <- subset_samples(ecm_phy, ecm_phy@sam_data$specification_value %in% c("UP", "LO")) %>% clean_pq(remove_empty_samples = T,remove_empty_taxa = T, clean_samples_names = T)

colnames(ecm_otusam)
ecm_otusam <- data.frame(ecm_phy@sam_data, t(ecm_phy@otu_table))
```

*VAR TRANSFO FOR ANALYSIS*
```{r transfo variables}
sam <- lapply(physeq@sam_data, as.factor) %>% as.data.frame()
sample_data(physeq)$year_factor <- as.factor(sample_data(physeq)$year)
```

```{r transposing otu table in physq object for hill_pq}
phy_otu_t <- t(physeq@otu_table)
phy_sam <- physeq@sam_data
phy_tax <- physeq@tax_table

ecm_phy_otu_t <- t(ecm_phy@otu_table) %>% otu_table()
ecm_phy_t <- phyloseq(ecm_phy_otu_t, ecm_phy@sam_data, ecm_phy@tax_table)
```

```{r grouping by factor of insterest}
yr_12 <- merge_group(ecm_12, group = "treatment_edge",)
yr_18 <- merge_group(ecm_18, group = "treatment_edge",)
```
